### 2016-09-06 Chapters/QuantumMechanics.tex
Chaps. 1, 2, and 5 are complete

It the language of second quantization, we assume that a, not
necessarily complete, set of orthonormal one-particle states is
available. These are usually called \glspl{MO}.
In general, we would like to embed as many of the symmetries of the
system into the one-particle basis. This is accomplished by requiring
the one-particle space to be the common eigenbasis for a complete set of
suitable, commuting one-particle operators.\autocite{Gross1991-hi}
Any $N$-particle state can be constructed by distributing particles
into one-particle states and accordingly labelling them by the number of
particles per one-particle states present.
This is the so-called \gls{ON} vector representation:
\begin{equation}
  \ket{\vect{n}} = \ket{n1,n2,\ldots},
\end{equation}
and the span of all \acrshort{ON} vectors is called the \emph{Fock
space} $F^N$, a space isomorphic to the Hilbert space $H^N$ of
$N$-particle wave functions.
To preserve consistency, the \emph{vacuum} state with no particles:
\begin{equation}
 \ket{\mathrm{vac}} = \ket{0,0,\ldots}
\end{equation}
is included in the construction of Fock space.
Any $N$-particle state can be generated by application of
\emph{creation} operators $\cons{i}$. Creation operators increase the
occupation number in the one-particle state $i$ by $1$, up to a phase.
Conversely, \emph{annihilation} operators $\anni{i}$, defined as the Hermitian
conjugates of the creation operators, decrease occupation in state $i$
by $1$, again up to a phase. Clearly, annihilation of any one-particle
state in the vacuum results yields $0$.
Pauli's exclusion principle is enforced by the \emph{canonical
anticommutation relations}:
\begin{subequations}
  \begin{align}
    \cons{r}\cons{s} + \cons{s}\cons{r} &= 0 \\
    \anni{r}\anni{s} + \anni{s}\anni{r} &= 0 \\
    \cons{r}\anni{s} + \anni{s}\cons{r} &= \delta_{rs},
  \end{align}
\end{subequations}
showing how the construction of the Fock space representation
embeds antisymmetry at the operator level and not in the wave functions.
Finally, the second-quantized molecular electronic Hamiltonian in the \acrlong{BO}
approximation is:\autocite{Helgaker2000-tz, Shavitt2009-mr}
\begin{equation}\label{eq:sq-hamiltonian}
  H = \sum_{rs}h_{rs}\cons{r}\anni{s}
  + \frac{1}{2}\sum_{rstu}g_{rstu}\cons{r}\cons{t}\anni{u}\anni{s}
  + V_\mathrm{NN}
\end{equation}
where the matrix elements are given as integrals over the chosen
one-particle basis:
\begin{subequations}
 \begin{align}
   h_{rs} &= \int \diff\vect{r} \varphi_r^\dagger(\vect{r}) h(\vect{r})\varphi_s(\vect{r}) \label{eq:one-el}\\
   g_{rstu} &=
   \int\diff\vect{r}\int\diff\vect{r}^\prime
   \Omega_{rs}(\vect{r}) g(\vect{r},\vect{r}^\prime) \Omega_{tu}(\vect{r}^\prime) \label{eq:two-el} \\
  \Omega_{rs}(\vect{r}) &= \varphi_r^\dagger(\vect{r})\varphi_s(\vect{r}). \label{eq:overlap}
 \end{align}
\end{subequations}
The second-quantized Hamiltonian Eq.~\eqref{eq:sq-hamiltonian} is thus a
\emph{projected} operator, exact only within the subspace determined by
the chosen one-particle set.~\autocite{Gross1991-hi, Helgaker2000-tz}

@@

The idea behind mean-field theory is simple: seek the best description
of the many-electron wave function using just one Slater determinant.
How to direct this search towards the "best"? We will use the
variational principle and optimize the degrees of freedom in our
single-determinantal trial wave function accordingly.
It is evident that there is no variational freedom in the
$N$-electron space. All variational degrees of freedom reside in
the one-electron space, \ie in the space of \glspl{MO}
$\lbrace\varphi_r\rbrace$, used to build the determinant $\Ket{0}$.
The one-electron \glspl{MO} are thus optimized considering each single
electron in the mean field generated by the motion of all other
electrons, hence the name of the method. Clearly this neglects
parts of the electron--electron interaction energy, the \emph{electron
correlation}. However, in most cases, it provides a suitable starting
approximation for more sophisticated many-body treatments that recover
larger portions of the correlation energy.

Starting from an initial guess for the one-particle states, we
iteratively refine them by applying suitable rotations.
The iterations are carried out until a suitable convergence criterion
has been met, \ie the orbitals are \emph{self-consistent}.
The unitary rotation operator can be expressed as the exponential of a
one-electron, anti-Hermitian operator:
\begin{alignat}{3}
  U = \expo{-\kappa},
  \quad&
  \kappa = \sum_{rs}
  [\kappa_{rs} \cons{r}\anni{s} - \kappa^*_{rs}\cons{s}\anni{r}],
  \quad&
  \kappa^\dagger = -\kappa.
\end{alignat}
The energy of the system is now a function of the rotation parameters
$\kappa_{rs}, \kappa^*_{rs}$ and setting its gradient $\mat{E}^{[1]}$ to
zero will locate a stationary point. The gradient of the electronic
energy can be identified by comparing the Taylor and \gls{BCH}
commutator expansions of the Hamiltonian expectation value.
For a closed-shell this yields \emph{Brillouin's theorem}:
\begin{equation}\label{eq:brillouin}
  E^{[1]}_{ai} = -f_{ai}
 = h_{ai} + \sum_{j}[g_{aijj} - g_{ajji}] = 0,
\end{equation}
with $\mat{F}$ being the \emph{Fock matrix}.
The two-electron terms appearing in the Fock matrix are called the
Coulomb and exchange integrals, respectively.
The stationarity condition implies that the Fock matrix is
block-diagonal in the basis of the optimal \glspl{MO}.
In the \emph{canonical representation} one seeks the set of orbitals
that make the Fock matrix diagonal. Hence the optimization problem is
equivalent to diagonalization of an effective one-electron operator:
\begin{equation}\label{eq:fock-eq}
 \mat{F}\vect{\varphi} = \vect{\varphi}\mat{\epsilon}
\end{equation}
where the eigenvalues of the Fock matrix are called \emph{orbital
energies}.

A different approach is taken in \gls{DFT}. The electron density is
taken as the central quantity in the theory, instead of the wave
function.
The Hohenberg--Kohn theorems\autocite{Hohenberg1964-wo, Eschrig2012-as}
and their relativistic counterparts\autocite{Rajagopal1973-ns,
Dreizler2012-ay} establish the energy as a \emph{functional} of the
density.
The \gls{KS} approach to \acrshort{DFT} takes us a step further by
assuming the density of the system to be equal to that of a fictitious,
noninteracting system once again described by a single Slater
determinant.\autocite{Kohn1965-hg}
We rewrite the energy as a sum of five terms:
\begin{equation}\label{eq:ks-energy}
 E[\rho] = T_s[\rho] + V_\mathrm{ext}[\rho] + J[\rho] + E_\mathrm{xc}[\rho] + V_\mathrm{NN}
\end{equation}
the first four of which are functionals of the density. The first term
$T_s[\rho]$ is the kinetic energy of the fictitious noninteracting
system.
The second and third terms in Eq. \eqref{eq:ks-energy} represent,
respectively, the classical interaction of the electrons with the
external potential (including the nuclear attraction potential) and the
classical Coulomb interaction of the density with itself:
\begin{subequations}
 \begin{align}
   V_\mathrm{ext}[\rho] &= \int\diff\vect{r} V_\mathrm{ext}(\vect{r})\rho(\vect{r}) \\
   J[\rho] &=
   \frac{1}{2}\int\diff\vect{r}\int\diff\vect{r}^\prime
   \rho(\vect{r})g(\vect{r}, \vect{r}^\prime)\rho(\vect{r}^\prime)
 \end{align}
\end{subequations}
The fourth term in Eq. \eqref{eq:ks-energy} is the \gls{XC} functional.
This term accounts for the nonclassical part of the two-electron
interaction: electron exchange and electron correlation.
Moreover, it also corrects for the self-interaction present in $J[\rho]$
and the error introduced in the evaluation of the kinetic energy.
The exact form of this functional is not known and a number of
parametrized approximations exist.\autocite{Koch2015-wq}
In the \gls{LDA}, the functional is expressed in terms of an
\emph{energy density} $e_\mathrm{xc}$ which is a local function of the
density:
\begin{equation}
 E^\mathrm{LDA}_\mathrm{xc}[\rho] = \int e_\mathrm{xc}(\rho(\vect{r}))\diff\vect{r}.
\end{equation}
In the \gls{GGA}, the energy density is a local function of both the
density and the norm of its gradient
$\zeta(\vect{r})=\vect{\nabla}\rho\cdot \vect{\nabla}\rho$:
\begin{equation}
 E^\mathrm{GGA}_\mathrm{xc}[\rho] = \int e_\mathrm{xc}(\rho(\vect{r}),\zeta(\vect{r}))\diff\vect{r}.
\end{equation}
Finally, in \emph{hybrid} functionals, some proportion of the
\acrshort{HF} exchange is included:
\begin{equation}
 E^\mathrm{hybrid}_\mathrm{xc}[\rho] = E^\mathrm{GGA}_\mathrm{xc}[\rho] + \gamma E^\mathrm{HF}_\mathrm{x}[\rho]
\end{equation}
The \acrshort{KS}-\acrshort{DFT} state is optimized by the same
iterative process described above.
Repeated diagonalizations of the \acrshort{KS} matrix are carried out
until self-consistency is reached.
The \acrshort{KS} matrix contains the one-electron \acrshort{XC}
potential in addition to the other terms contained in the Fock matrix:
\begin{equation}\label{eq:ksmatrix}
  f_{rs} = h_{rs} + \sum_{j}[g_{rsjj} - \gamma g_{rjjs}] + v_{\mathrm{xc};rs}
\end{equation}

@@

As shown by \citeauthor{Koch1997-nm}, the \acrshort*{MBPT} series can be
derived from a \acrshort*{CC} ansatz for the wave function.
This ensures term-by-term size-extensivity of the energy at all
orders.\autocite{Helgaker2000-tz}
The governing equations
\eqref{eq:cc-amplitudes}--\eqref{eq:cc-multipliers} are expanded in
terms of the fluctuation potential. Energy corrections are identified
from the corresponding order expansion of the \acrshort*{CC} Lagrangian
and retaining only terms complying to the $2n+1$ and $2n+2$ rules.
A superscript index in square brackets - $[i]$ -  will denote an $i$-th
order contribution.

@@

To first order, one has:
\begin{equation}\label{eq:1st-order}
  \begin{aligned}
  \denom{\mu_2}\tamp{\mu_2}^{[1]}
  &= - \braket{\mu_2 | \Phi | \mathrm{HF}}
  = - \braket{\mathrm{HF} | \commutator{\Phi}{\cluster{\mu_2}} |
  \mathrm{HF}} \\
  &= \denom{\mu_2}\tbar{\mu_2}^{[1]}
  \end{aligned}
\end{equation}
which shows that singles, due to Brillouin's condition, and triples
amplitudes are of second and higher order.
To second order, one obtains:
\begin{equation}\label{eq:2nd-order}
  \begin{aligned}
  \denom{\mu_i}\tamp{\mu_i}^{[2]}
  &= - \braket{\mu_i | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = - \braket{\tbar{2}^{[1]} | \commutator{\Phi}{\cluster{\mu_i}} |
  \mathrm{HF}} \\
  &= \denom{\mu_i}\tbar{\mu_i}^{[2]},\quad i = 1, 2, 3.
  \end{aligned}
\end{equation}
According to the $2n+1$ and $2n+2$ rules, energy corrections up to fifth
order can be obtained from amplitudes and multipliers up to second
order.
\todo[inline]{CHECK THE LOGIC OF THIS STATEMENT}
Since the triples first appear to second order the \gls{CCSD}
energy is correct to third order in the fluctuation potential.
The triples contributions to the fourth and fifth order corrections to
the energy are given as:
\begin{subequations}
  \begin{align}
    E_T^{[4]} &=
 \braket{\tbar{2}^{[1]}| \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
 \label{eq:4th-order-triples} \\
    E_T^{[5]} &=
    \braket{\tbar{1}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{2}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
  &+ \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_2^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
    &+
    \frac{1}{2}\braket{\tbar{3}^{[2]} |
    \commutator{\commutator{\Phi}{T_2^{[1]}}}{T_2^{[1]}}
    | \mathrm{HF}}.
 \label{eq:5th-order-triples}
  \end{align}
\end{subequations}
These expressions form the basis for the development of reduced-scaling
\acrshort*{CC} schemes were the double and triple excitation are treated
approximately. These are the subject of the next Section.

>>>>

It the language of second quantization, we assume that a, not
necessarily complete, set of orthonormal one-particle states is
available. These are usually called \glspl{MO}.
In general, we would like to embed as many of the symmetries of the
system into the one-particle basis. This is accomplished by requiring
the one-particle space to be the common eigenbasis for a complete set of
suitable, commuting one-particle operators.\autocite{Gross1991-hi}
Any $N$-particle state can be constructed by distributing particles
into one-particle states and accordingly labelling them by the number of
particles per one-particle states present.
This is the so-called \gls{ON} vector representation:
\begin{equation}
  \ket{\vect{n}} = \ket{n1,n2,\ldots},
\end{equation}
and the span of all \acrshort{ON} vectors is called the \emph{Fock
space} $F^N$, a space isomorphic to the Hilbert space $H^N$ of
$N$-particle wave functions.
To preserve consistency, the \emph{vacuum} state with no particles:
\begin{equation}
 \ket{\mathrm{vac}} = \ket{0,0,\ldots}
\end{equation}
is included in the construction of Fock space.
Any $N$-particle state can be generated by application of
\emph{creation} operators $\cons{i}$. Creation operators increase the
occupation number in the one-particle state $i$ by $1$, up to a phase.
Conversely, \emph{annihilation} operators $\anni{i}$, defined as the Hermitian
conjugates of the creation operators, decrease occupation in state $i$
by $1$, again up to a phase. Clearly, annihilation of any one-particle
state in the vacuum results yields $0$.
Pauli's exclusion principle is enforced by the \emph{canonical
anticommutation relations}:
\begin{subequations}
  \begin{align}
    \cons{r}\cons{s} + \cons{s}\cons{r} &= 0 \\
    \anni{r}\anni{s} + \anni{s}\anni{r} &= 0 \\
    \cons{r}\anni{s} + \anni{s}\cons{r} &= \delta_{rs},
  \end{align}
\end{subequations}
showing how the construction of the Fock space representation
embeds antisymmetry at the operator level and not in the wave functions.
Finally, the second-quantized molecular electronic Hamiltonian in the \acrlong{BO}
approximation is:\autocite{Helgaker2000-tz, Shavitt2009-mr}
\begin{equation}\label{eq:sq-hamiltonian}
  H = \sum_{rs}h_{rs}\cons{r}\anni{s}
  + \frac{1}{2}\sum_{rstu}g_{rstu}\cons{r}\cons{t}\anni{u}\anni{s}
  + V_\mathrm{NN}
\end{equation}
where the matrix elements are given as integrals over the chosen
one-particle basis:
\begin{subequations}
 \begin{align}
   h_{rs} &= \int \diff\vect{r} \phi_r^\dagger(\vect{r}) h(\vect{r})\phi_s(\vect{r}) \label{eq:one-el}\\
   g_{rstu} &=
   \int\diff\vect{r}\int\diff\vect{r}^\prime
   \Omega_{rs}(\vect{r}) g(\vect{r},\vect{r}^\prime) \Omega_{tu}(\vect{r}^\prime) \label{eq:two-el} \\
  \Omega_{rs}(\vect{r}) &= \phi_r^\dagger(\vect{r})\phi_s(\vect{r}). \label{eq:overlap}
 \end{align}
\end{subequations}
The second-quantized Hamiltonian Eq.~\eqref{eq:sq-hamiltonian} is thus a
\emph{projected} operator, exact only within the subspace determined by
the chosen one-particle set.~\autocite{Gross1991-hi, Helgaker2000-tz}

@@

The idea behind mean-field theory is simple: seek the best description
of the many-electron wave function using just one Slater determinant.
How to direct this search towards the "best"? We will use the
variational principle and optimize the degrees of freedom in our
single-determinantal trial wave function accordingly.
It is evident that there is no variational freedom in the
$N$-electron space. All variational degrees of freedom reside in
the one-electron space, \ie in the space of \glspl{MO}
$\lbrace\phi_r\rbrace$, used to build the determinant $\Ket{0}$.
The one-electron \glspl{MO} are thus optimized considering each single
electron in the mean field generated by the motion of all other
electrons, hence the name of the method. Clearly this neglects
parts of the electron--electron interaction energy, the \emph{electron
correlation}. However, in most cases, it provides a suitable starting
approximation for more sophisticated many-body treatments that recover
larger portions of the correlation energy.

Starting from an initial guess for the one-particle states, we
iteratively refine them by applying suitable rotations.
The iterations are carried out until a suitable convergence criterion
has been met, \ie the orbitals are \emph{self-consistent}.
The unitary rotation operator can be expressed as the exponential of a
one-electron, anti-Hermitian operator:
\begin{alignat}{3}
  U = \expo{-\kappa},
  \quad&
  \kappa = \sum_{rs}
  [\kappa_{rs} \cons{r}\anni{s} - \kappa^*_{rs}\cons{s}\anni{r}],
  \quad&
  \kappa^\dagger = -\kappa.
\end{alignat}
The energy of the system is now a function of the rotation parameters
$\kappa_{rs}, \kappa^*_{rs}$ and setting its gradient $\mat{E}^{[1]}$ to
zero will locate a stationary point. The gradient of the electronic
energy can be identified by comparing the Taylor and \gls{BCH}
commutator expansions of the Hamiltonian expectation value.
For a closed-shell this yields \emph{Brillouin's theorem}:
\begin{equation}\label{eq:brillouin}
  E^{[1]}_{ai} = -f_{ai}
 = h_{ai} + \sum_{j}[g_{aijj} - g_{ajji}] = 0,
\end{equation}
with $\mat{F}$ being the \emph{Fock matrix}.
The two-electron terms appearing in the Fock matrix are called the
Coulomb and exchange integrals, respectively.
The stationarity condition implies that the Fock matrix is
block-diagonal in the basis of the optimal \glspl{MO}.
In the \emph{canonical representation} one seeks the set of orbitals
that make the Fock matrix diagonal. Hence the optimization problem is
equivalent to diagonalization of an effective one-electron operator:
\begin{equation}\label{eq:fock-eq}
 \mat{F}\vect{\phi} = \vect{\phi}\mat{\epsilon}
\end{equation}
where the eigenvalues of the Fock matrix are called \emph{orbital
energies}.

A different approach is taken in \gls{DFT}, where the electron density
is the central quantity in the theory, instead of the wave function.
The Hohenberg--Kohn theorems\autocite{Hohenberg1964-wo, Eschrig2012-as}
and their relativistic counterparts\autocite{Rajagopal1973-ns,
Dreizler2012-ay} establish the energy as a \emph{functional} of the
density.
The \gls{KS} approach to \acrshort{DFT} takes us a step further by
assuming the density of the system to be equal to that of a fictitious,
noninteracting system once again described by a single Slater
determinant.\autocite{Kohn1965-hg}
We rewrite the energy as a sum of five terms:
\begin{equation}\label{eq:ks-energy}
 E[\rho] = T_s[\rho] + V_\mathrm{ext}[\rho] + J[\rho] + E_\mathrm{xc}[\rho] + V_\mathrm{NN}
\end{equation}
the first four of which are functionals of the density. The first term
$T_s[\rho]$ is the kinetic energy of the fictitious noninteracting
system.
The second and third terms in Eq. \eqref{eq:ks-energy} represent,
respectively, the classical interaction of the electrons with the
external potential (including the nuclear attraction potential) and the
classical Coulomb interaction of the density with itself:
\begin{subequations}
 \begin{align}
   V_\mathrm{ext}[\rho] &= \int\diff\vect{r} V_\mathrm{ext}(\vect{r})\rho(\vect{r}) \\
   J[\rho] &=
   \frac{1}{2}\int\diff\vect{r}\int\diff\vect{r}^\prime
   \rho(\vect{r})g(\vect{r}, \vect{r}^\prime)\rho(\vect{r}^\prime)
 \end{align}
\end{subequations}
The fourth term in Eq. \eqref{eq:ks-energy} is the \gls{XC} functional.
This term accounts for the nonclassical part of the two-electron
interaction: electron exchange and electron correlation.
Moreover, it also corrects for the self-interaction present in $J[\rho]$
and the error introduced in the evaluation of the kinetic energy.
The exact form of this functional is not known and a number of
parametrized approximations exist.\autocite{Koch2015-wq}
In the \gls{LDA}, the functional is expressed in terms of an
\emph{energy density} $e_\mathrm{xc}$ which is a local function of the
density:
\begin{equation}
 E^\mathrm{LDA}_\mathrm{xc}[\rho] = \int e_\mathrm{xc}(\rho(\vect{r}))\diff\vect{r}.
\end{equation}
In the \gls{GGA}, the energy density is a local function of both the
density and the norm of its gradient
$\zeta(\vect{r})=\vect{\nabla}\rho\cdot \vect{\nabla}\rho$:
\begin{equation}
 E^\mathrm{GGA}_\mathrm{xc}[\rho] = \int e_\mathrm{xc}(\rho(\vect{r}),\zeta(\vect{r}))\diff\vect{r}.
\end{equation}
Finally, in \emph{hybrid} functionals, some proportion of the
\acrshort{HF} exchange is included:
\begin{equation}
 E^\mathrm{hybrid}_\mathrm{xc}[\rho] = E^\mathrm{GGA}_\mathrm{xc}[\rho] + \gamma E^\mathrm{HF}_\mathrm{x}[\rho]
\end{equation}
The \acrshort{KS}-\acrshort{DFT} state is optimized by the same
iterative process described above.
Repeated diagonalizations of the \acrshort{KS} matrix are carried out
until self-consistency is reached.
The \acrshort{KS} matrix:
\begin{equation}\label{eq:ksmatrix}
  f_{rs} = h_{rs} + \sum_{j}[g_{rsjj} - \gamma g_{rjjs}] + f_{\mathrm{xc};rs}.
\end{equation}
contains an \acrshort{XC} one-electron contribution due to the
in addition to the other terms contained in the Fock matrix.
The functional derivative of the \acrshort*{XC} functional enters in the
definition of this additional term:
\begin{alignat}{2}
  f_{\mathrm{xc}; rs}
  =
  \int\diff\vect{r}\Omega_{rs}(\vect{r})v_\mathrm{xc}(\vect{r}),
  \quad&
  v_\mathrm{xc}(\vect{r})
  \coloneq
  \frac{\delta E_\mathrm{xc}[\rho(\vect{r})]}{\delta \rho(\vect{r})}
\end{alignat}
notice moreover the exact exchange admixing factor $\gamma$ in the
two-electron part of the \acrshort*{KS} matrix.

How are \glspl*{MO} constructed? The usual approach is to expand them in
a basis of one-particle, atom-centered basis functions of known
functional form:
\begin{equation}
  \phi_r(\vect{r})
  =
  \sum_{\alpha=1}^{N_\mathrm{basis}}
  C_{\alpha r}\chi_{\alpha}(\vect{r})
\end{equation}
and optimize the expansion coefficients in the \acrshort*{SCF} procedure.
Choosing the one-particle \acrshort*{MO} space is then equivalent to
choosing the one-particle \gls{AO} space.
The functional form of the \acrshort*{AO} basis is chosen to strike a
balance between a physically accurate description of the short-range
Coulomb interaction and algorithmic robustness and
efficiency.\autocite{Helgaker2000-tz}
The usual choice, also adopted in this thesis, is to use bases of
Gaussian functions.\autocite{Helgaker1995-gi, Reine2012-tq}

@@

As shown by \citeauthor{Koch1997-nm}, the \acrshort*{MBPT} series can be
derived from a \acrshort*{CC} ansatz for the wave function.
This ensures term-by-term size-extensivity of the energy at all
orders.\autocite{Helgaker2000-tz}
The governing equations
\eqref{eq:cc-amplitudes}--\eqref{eq:cc-multipliers} are expanded in
terms of the fluctuation potential. Energy corrections are identified
from the corresponding order expansion of the \acrshort*{CC} Lagrangian
and retaining only terms complying to the $2n+1$ and $2n+2$ rules.
\autocite{Helgaker1988-to, Helgaker1989-wl, Helgaker1992-ph}
A superscript index in square brackets - $[i]$ -  will denote an $i$-th
order contribution.

@@

To first order, one has:
\begin{equation}\label{eq:1st-order}
  \begin{aligned}
  \denom{\mu_2}\tamp{\mu_2}^{[1]}
  &= - \braket{\mu_2 | \Phi | \mathrm{HF}}
  = - \braket{\mathrm{HF} | \commutator{\Phi}{\cluster{\mu_2}} |
  \mathrm{HF}} \\
  &= \denom{\mu_2}\tbar{\mu_2}^{[1]}
  \end{aligned}
\end{equation}
which shows that singles, due to Brillouin's condition, and triples
amplitudes are of second and higher order.
To second order, one obtains:
\begin{equation}\label{eq:2nd-order}
  \begin{aligned}
  \denom{\mu_i}\tamp{\mu_i}^{[2]}
  &= - \braket{\mu_i | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = - \braket{\tbar{2}^{[1]} | \commutator{\Phi}{\cluster{\mu_i}} |
  \mathrm{HF}} \\
  &= \denom{\mu_i}\tbar{\mu_i}^{[2]},\quad i = 1, 2, 3.
  \end{aligned}
\end{equation}
According to the $2n+1$ and $2n+2$ rules, energy corrections up to fifth
order can be obtained from amplitudes and multipliers up to second
order. Since the triples first appear to second order, the \gls{CCSD}
energy is correct to third order in the fluctuation potential.
The triples contributions to the fourth and fifth order corrections to
the energy are given as:
\begin{subequations}
  \begin{align}
    E_T^{[4]} &=
 \braket{\tbar{2}^{[1]}| \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
 \label{eq:4th-order-triples} \\
    E_T^{[5]} &=
    \braket{\tbar{1}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{2}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
  &+ \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_2^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
    &+
    \frac{1}{2}\braket{\tbar{3}^{[2]} |
    \commutator{\commutator{\Phi}{T_2^{[1]}}}{T_2^{[1]}}
    | \mathrm{HF}}.
 \label{eq:5th-order-triples}
  \end{align}
\end{subequations}
These expressions form the basis for the development of reduced-scaling
\acrshort*{CC} schemes were the double and triple excitation are treated
approximately. These are the subject of the next Section.

### 2016-09-06 Chapters/ResponseTheory.tex
Work on MQM chapter
Minor adjustments to text
Chaps. 1, 2, and 5 are complete

\epigraph{
White light goin' messin' up my brain

White light it's gonna drive me insane
}{
--- \textsc{The Velvet Underground}
}
\epigraph{What's the frequency, Kenneth? [...]

          I never understood the frequency, uh-huh}{
          --- \textsc{R.E.M.}}

\begin{itemize}
    \item Something about exact state response theory: quasienergy,
      response functions, residues of response functions.
    \item  Derivation of the linear response function with
      quantum/classical polarizable Hamiltonian and then refer to the
      OpenRSP paper for the open-ended treatment.~\autocite{pcm-openrsp}
\end{itemize}

\pagebreak

@@

General theory~\autocite{Olsen1985-nr, Christiansen1998-pe,
Norman2011-ad, Helgaker2012-cz, Pawlowski2015-sq}
Solution strategy for response equations~\autocite{Kauczor2011-rd}
Open-ended \acrshort{AO}-based \acrshort{SCF} response~\autocite{
Larsen2000-hj, Kjaergaard2008-hy, Thorvaldsen2008-sg,
Kristensen2008-hv, Ringholm2014-gx}
Open-ended \acrshort{SCF} residues~\autocite{Friese2015-kb}

@@

\acrshort{PCM}~\autocite{Cammi1994-qj, Cammi1996-wf, Cammi1996-vx,
Cammi1999-rb, Cammi2003-qy, Frediani2005-nc, Ferrighi2010-pm}
\acrshort{FQ} and \acrshort{FQ}/\acrshort{PCM}~\autocite{Lipparini2012-hx, Lipparini2012-tl, Lipparini2013-ud}
\acrshort{PE}~\autocite{Olsen2010-wa}
\acrshort{QM}/\acrshort{MM}/Continuum~\autocite{Stenidal2011-ki, Caprasecca2012-ir}
Paper V~\autocite{pcm-openrsp}

>>>>

\begin{epigraphs}
\qitem{
White light goin' messin' up my brain

White light it's gonna drive me insane
}{
--- \textsc{The Velvet Underground}
}
\qitem{What's the frequency, Kenneth? [...]

          I never understood the frequency, uh-huh}{
          --- \textsc{R.E.M.}}
\end{epigraphs}

The scope of quantum chemistry is to understand and predict molecular
properties, but so far we have made no mention of properties other than
the energy.
The experimentalists' view of molecular systems is built mainly around the use
of spectroscopic techniques that explore the interaction of light and matter.
When a system is exposed to an external perturbing electromagnetic field, it will
respond with a detectable change in its properties.\autocite{Pedersen2012-il, Jaszunski2012-wy}
By "external" we mean an electromagnetic field which is "weak" when
compared to the electron--electron and electron--nuclei interactions in
the molecule.
\emph{Response theory} is the missing link between theory and
experiment. It allows to describe perturbation-induced changes in
molecular properties in terms of \emph{response functions}, built solely
by means of \emph{unperturbed} molecular states and energies.
Moreover, excitation energies and transition moments can be inferred
from response functions without recurring to an explicit modelling of
excited states.
Section \ref{sec:exact-response} will give a brief introduction to exact
state response theory for isolated molecules. The concepts of
quasienergy,\autocite{Christiansen1998-pe} variational perturbation
theory\autocite{Helgaker1992-ph} and pole-and-residue
analysis of response functions\autocite{Olsen1985-nr} will be introduced.
I will explicitly derive the \acrshort{SCF} response function for a
quantum/classical polarizable Hamiltonian in Section
\ref{sec:csm-response}. The derivation will employ the
density matrix-based, \acrshort{AO} formalism introduced by
\citeauthor{Thorvaldsen2008-sg} in \noparcite[ref.][]{Thorvaldsen2008-sg}.
This formalism is suitable for the derivation of arbitrary-order
response functions and their residues, also when perturbation-dependent
basis sets are considered.\autocite{Friese2015-kb, Friese2015-bu}
Such a general derivation for the \acrshort{PCM} is presented in
\paper{V}.

The response treatment of molecular properties has its roots in
time-dependent perturbation theory\autocite{Konishi2009-zb} and has been
continually developed in quantum chemistry for the past 30 years:
the exposition will be necessarily brief and incomplete, the reader is
referred to the cited works in the literature for a more detailed
presentation of the topic.\autocite{Olsen1985-nr, Helgaker1992-ph, Olsen1995-pf,
Christiansen1998-pe, Norman2011-ad, Helgaker2012-cz, Pawlowski2015-sq}

@@

General theory~\autocite{Olsen1985-nr, Helgaker1992-ph, Olsen1995-pf,
Christiansen1998-pe, Norman2011-ad, Helgaker2012-cz, Pawlowski2015-sq}
Solution strategy for response equations~\autocite{Saad2003-oa,
Saad2011-gm, Kauczor2011-rd, Malmqvist2013-vw}
Open-ended \acrshort{AO}-based \acrshort{SCF} response~\autocite{
Larsen2000-hj, Kjaergaard2008-hy, Thorvaldsen2008-sg,
Kristensen2008-hv, Ringholm2014-gx}
Open-ended \acrshort{SCF} residues~\autocite{Friese2015-kb}

@@

\acrshort{PCM}~\autocite{Cammi1994-qj, Cammi1996-wf, Cammi1996-vx,
Cammi1999-rb, Cammi2003-qy, Frediani2005-nc, Ferrighi2010-pm}
\acrshort{FQ} and \acrshort{FQ}/\acrshort{PCM}~\autocite{Lipparini2012-hx, Lipparini2012-tl, Lipparini2013-ud}
\acrshort{PE}~\autocite{Olsen2010-wa}
\acrshort{QM}/\acrshort{MM}/Continuum~\autocite{Steindal2011-ki, Caprasecca2012-ir}
Paper V~\autocite{pcm-openrsp}
